Non-word repetition in children learning Yélî Dnye

### 2021-08-20 manuscript.tex
first stab at R1

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[english,,man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdftitle={Non-word repetition in children learning Yélî Dnye},
            pdfkeywords={phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

@@

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\makeatother
\shorttitle{NWR in Yélî Dnye learners}
\author{Alejandrina Cristia\textsuperscript{1}\ \& Marisa Casillas\textsuperscript{2,3}}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes Cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics\\\textsuperscript{3} University of Chicago}
\authornote{

Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d’Ulm, 75005 Paris, France. E-mail: alecristia@gmail.com}
\keywords{phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness\newline\indent Word count: ~11,500 words}
\usepackage{lineno}

@@

Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood (e.g., Hazan \& Barrett, 2000). Much of the evidence for later phonological development comes from non-word repetition (NWR) tasks. In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning. After hearing this non-word, the participant's task is to try to immediately and precisely repeat it. NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are ({\textbf{???}}), and how extensively phonological constraints found in the lexicon affect online production (Gallagher, 2014). NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders (Estes, Evans, \& Else-Quest, 2007). Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe (COST Action, 2009; Meir, Walters, \& Armon-Lotem, 2016). NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before). In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG), which has a large and unusually dense phonological inventory. The study was designed to contribute to four broad research areas, three via direct results.

@@

Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages. Some work documents much lower NWR scores for longer, compared to shorter, items (e.g., among Cantonese-learning children; {\textbf{???}}), whereas differences are negligible in other studies (e.g., among Italian learners; Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019). It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture. In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer. To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths. We found data for learners of Israeli Arabic (Jaber-Awida, 2018); Cantonese ({\textbf{???}}); English (Vance, Stackhouse, \& Wells, 2005); Italian (Piazzalunga et al., 2019); and Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020); and integrated those data with Yélî Dnye results from the present study in Figure \ref{fig:fig-prevlit}.

@@

The third research area we contribute data to relates to the possibility that individual variation in NWR scores is structured. Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based ({\textbf{???}}). In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size. For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education. Based on previous work, we looked at potential increases with age ({\textbf{???}}; Farmani et al., 2018; Vance et al., 2005). Prior research typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; {\textbf{???}}; Balladares, Marshall, \& Griffiths, 2016; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007). Although past research has not investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (e.g., Havron et al., 2019), and therefore we report on these too.

@@

Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, and always meeting the condition that they do not mean anything in the target language (e.g., Balladares et al., 2016; Wilsenach, 2013). We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye. We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat (other work has also used up to 24-30 items: Jaber-Awida, 2018; {\textbf{???}}), with the entire test inventory of 40 final items distributed across children.

@@

In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several of the studies in Figure \ref{fig:fig-prevlit} (English Vance et al., 2005; Italian Piazzalunga et al., 2019; Cantonese {\textbf{???}}; but note Cristia et al., 2020 is an exception). Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency. In brief, our results underscore the idea that phonological development continues well past the first few years of life, extending into middle childhood and perhaps later (Hazan \& Barrett, 2000).

In contrast, previous work shows little evidence for effects of maternal education (e.g., Farmani et al., 2018; {\textbf{???}}; Meir \& Armon-Lotem, 2017) on NWR scores. We did not expect large effects of maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally. In fact, maternal education correlated with NWR score at about r\textasciitilde{}.1, which is small. Similarly, NWR scores may not vary greatly with participant gender according to previous work (Chiat \& Roy, 2007), and for that as well we find effects of about that size.

@@

To enrich this discussion, we looked for previous studies on monolingual children with normative development learning diverse languages, and entered them when they reported non-word repetition scores based on whole item scoring. We entered data from 14 studies (including ours), presenting data from 12 languages. Specifically, Arabic was represented by Jaber-Awida (2018); Cantonese by ({\textbf{???}}); English by Vance et al. (2005); Italian by Piazzalunga et al. (2019); Mandarin by Lei et al. (2011); Persian by Farmani et al. (2018); Slovak by Kapalková, Polišenská, and Vicenová (2013) and Polišenská and Kapalková (2014); Sotho by Wilsenach (2013); Spanish by Balladares et al. (2016); Swedish by ({\textbf{???}}) and Radeborg, Barthelom, SjöBerg, and Sahlén (2006); Tsimane' by Cristia et al. (2020); and Yélî Dnye from the present study. Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability. Results separating different age groups are shown in Figure \ref{fig:fig-prevlit-overall}.

@@

@@

@@

>>>>

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[english,,man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdftitle={Non-word repetition in children learning Yélî Dnye},
            pdfkeywords={phonology, non-word repetition, Papuan, non-industrial, non-urban, comparative, typology, markedness, literacy},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

@@

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\makeatother
\shorttitle{NWR in Yélî Dnye learners}
\author{Alejandrina Cristia\textsuperscript{1}\ \& Marisa Casillas\textsuperscript{2,3}}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes Cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics\\\textsuperscript{3} University of Chicago}
\authornote{Both authors contributed to study funding, design, data collection, annotation, analyses, writing.

Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d’Ulm, 75005 Paris, France. E-mail: alecristia@gmail.com}
\keywords{phonology, non-word repetition, Papuan, non-industrial, non-urban, comparative, typology, markedness, literacy\newline\indent Word count: ~11,500 words}
\usepackage{lineno}

@@

Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood (e.g., Hazan \& Barrett, 2000). Much of the evidence for later phonological development comes from non-word repetition (NWR) tasks. In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning. After hearing this non-word, the participant's task is to try to immediately and precisely repeat it. NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are (Bowey, 2001), and how extensively phonological constraints found in the lexicon affect online production (Gallagher, 2014). NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders (Estes, Evans, \& Else-Quest, 2007). Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe (COST Action, 2009; Meir, Walters, \& Armon-Lotem, 2016). NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before). In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG), which has a large and unusually dense phonological inventory. The study was designed to contribute to four broad research areas, three via direct results.

@@

Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages. Some work documents much lower NWR scores for longer, compared to shorter, items (e.g., among Cantonese-learning children; Stokes, Wong, Fletcher, \& Leonard, 2006), whereas differences are negligible in other studies (e.g., among Italian learners; Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019). It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture. In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer. To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths. We found data for learners of Israeli Arabic (Jaber-Awida, 2018); Cantonese (Stokes et al., 2006); English (Vance, Stackhouse, \& Wells, 2005); Italian (Piazzalunga et al., 2019); and Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020); and integrated those data with Yélî Dnye results from the present study in Figure \ref{fig:fig-prevlit}.

@@

The third research area we contribute data to relates to the possibility that individual variation in NWR scores is structured. Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based (Farabolini, Rinaldi, Caselli, \& Cristia, 2021). In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size. For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education. Based on previous work, we looked at potential increases with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance et al., 2005). Prior research typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; Balladares, Marshall, \& Griffiths, 2016; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007). Although past research has not investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (e.g., Havron et al., 2019), and therefore we report on these too.

@@

Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, and always meeting the condition that they do not mean anything in the target language (e.g., Balladares et al., 2016; Wilsenach, 2013). We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye. We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat (other work has also used up to 24-30 items: Jaber-Awida, 2018; Kalnak et al., 2014), with the entire test inventory of 40 final items distributed across children.

@@

In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several of the studies in Figure \ref{fig:fig-prevlit} (English Vance et al., 2005; Italian Piazzalunga et al., 2019; Cantonese Stokes et al., 2006; but note Cristia et al., 2020 is an exception). Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency. In brief, our results underscore the idea that phonological development continues well past the first few years of life, extending into middle childhood and perhaps later (Hazan \& Barrett, 2000).

In contrast, previous work shows little evidence for effects of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) on NWR scores. We did not expect large effects of maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally. In fact, maternal education correlated with NWR score at about r\textasciitilde{}.1, which is small. Similarly, NWR scores may not vary greatly with participant gender according to previous work (Chiat \& Roy, 2007), and for that as well we find effects of about that size.

@@

To enrich this discussion, we looked for previous studies on monolingual children with normative development learning diverse languages, and entered them when they reported non-word repetition scores based on whole item scoring. We entered data from 14 studies (including ours), presenting data from 12 languages. Specifically, Arabic was represented by Jaber-Awida (2018); Cantonese by Stokes et al. (2006); English by Vance et al. (2005); Italian by Piazzalunga et al. (2019); Mandarin by Lei et al. (2011); Persian by Farmani et al. (2018); Slovak by Kapalková, Polišenská, and Vicenová (2013) and Polišenská and Kapalková (2014); Sotho by Wilsenach (2013); Spanish by Balladares et al. (2016); Swedish by Kalnak et al. (2014) and Radeborg, Barthelom, SjöBerg, and Sahlén (2006); Tsimane' by Cristia et al. (2020); and Yélî Dnye from the present study. Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability. Results separating different age groups are shown in Figure \ref{fig:fig-prevlit-overall}.

@@

\hypertarget{data-code-and-materials-availability-statement}{%
\subsection{Data, code and materials availability statement}\label{data-code-and-materials-availability-statement}}

All data, code, and materials are available from \url{https://osf.io/5qspb/}

@@

\leavevmode\hypertarget{ref-bowey2001nonword}{}%
Bowey, J. A. (2001). Nonword repetition and young children's receptive vocabulary: A longitudinal study. \emph{Applied Psycholinguistics}, \emph{22}(3), 441--469.

@@

\leavevmode\hypertarget{ref-kalnak2014nonword}{}%
Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B. (2014). Nonword repetition--a clinical marker for specific language impairment in Swedish associated with parents' language-related problems. \emph{PloS One}, \emph{9}(2), e89544.
